---
title: "Data visualisation with ggplot"
output: github_document
author: Ludovic Dutoit, summary of @carpentries.org content
---

The code below aims to be a very short refresher and summarises what we have done together in class.

The whole lesson is well-written and available at:
[https://datacarpentry.org/R-ecology-lesson/04-visualization-ggplot2.html](https://datacarpentry.org/R-ecology-lesson/04-visualization-ggplot2.html)

The [R graph gallery](R graph gallery) also contains loads of creative and clean plots with their associated code


## A simple plot
loading the data and the library. The data is species observation on different type of field plots in different years.
```{r,message=F}
library("tidyverse")
surveys_complete <- read_csv("data/surveys_complete.csv")
```

The essence of the ggplot is of the following format.

```
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +  <GEOM_FUNCTION>()
```

data and mapping will inform ggplot on what we try to plot on the x and y axis, but without GEOM_FUNCTION, no point, line or fancy histograms.

```{r}
ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length))
```

Let's plot actual points by adding our first geom, geom_point()

```{r}
ggplot( data = surveys_complete, mapping = aes(x= weight, y = hindfoot_length)) + geom_point()
```


The whole core is a bit tedious to write, so we can save it into an object to be able to plot quickly the same data, varying all the geom.

```{r}
surveys_plot <-ggplot(data = surveys_complete, mapping = aes(x = weight, y = hindfoot_length))
surveys_plot + geom_point()
```


Let's use some blue points with some transparency

```
surveys_plot  + geom_point(alpha = 0.1, col = "blue")
```

Let's color by species_id
```{r}
surveys_plot + geom_point(alpha = 0.1, aes(color = species_id))
```

## Boxplot
We will now plot weight against species_id, one boxplot per species.
Let's save the data and mapping into an object:

```{r}
boxplot_weight <- ggplot(data = surveys_complete, mapping = aes( x= species_id, y=weight)) 
boxplot_weight
```
Now we can add the actual_boxplot and other geom_functions:

```{r}
boxplot_weight + geom_boxplot()
```

and with jitter, two plots on one just by adding an extra +

```{r}
boxplot_weight +  geom_boxplot() + geom_jitter(alpha=0.3, color= "tomato")
```

## Time series

Let's create a plot of genus through time. We can use dplyr and the pipe ( %>%) to count the number of occurences of each genus year by year.

```{r}
yearly_counts <- surveys_complete %>%
  count(year, genus)
```

Let's have a quick look:
```{r}
head(yearly_counts)
```

It seem to work.

Let's try a new geom, geom_line() to plot lines.
```{r}
ggplot(data = yearly_counts, mapping = aes(x= year, y = n)) +
  geom_line()
```

Oops, all the genera together does not work well, let's split and have one line per genus. That can be easily done in ggplot():

```{r}
ggplot(data = yearly_counts, aes(x= year, y = n, group = genus)) + geom_line()
```

let's add a color key:
```{r}
ggplot(data = yearly_counts, aes(x = year, y =n, color = genus)) + geom_line()
```

## Faceting
Let's use **faceting** and add the facet_wrap which allows us to split plots into subplots using a specific variable:
```{r}
ggplot(data = yearly_counts, aes(x= year, y =n )) + geom_line() + facet_wrap(facets = vars(genus))
```


Let's plot each sex for each year in each genus. Group_by() will group observation by year, genus and then sex before being counted using tally().

```{r}
yearly_sex_counts <- surveys_complete %>% 
  group_by(year, genus, sex)  %>% 
  tally()
head(yearly_sex_counts)
```

Now let's plot it, notice how we pass sex in the color argument and create subplots using the facet_wrap() function.

```{r}
ggplot(data = yearly_sex_counts, mapping = aes( x = year, y=n, color= sex))+ geom_line() + facet_wrap(facets = vars(genus) ) 
```

One can also organise the facets a bit more specifically using facet_grid:
An example with one line male, one line female

```{r}
ggplot(data = yearly_sex_counts, mapping = aes(x= year, y=n, color=sex)) + geom_line() + facet_grid(rows=vars(sex), cols = vars(genus)) 
```


## A little bit more customisation

There is a lot more plotting that can be done using iterative building to add functions one after the other separated by the + sign.


One can add [themes](https://ggplot2.tidyverse.org/reference/ggtheme.html) to change the general layout.
```{r}
ggplot(data = yearly_sex_counts, mapping = aes( x = year, y=n, color= sex))+ geom_line() + facet_wrap(facets = vars(genus) ) + theme_bw() 
```

For more content, I highly encourage you to refer to the [original lesson](https://datacarpentry.org/R-ecology-lesson/04-visualization-ggplot2.html). 


Let's play with vertical axes, and title for axes as well. 

```{r}
ggplot(data = yearly_sex_counts, mapping = aes( x = year, y=n, color= sex))+ geom_line() + facet_wrap(facets = vars(genus) ) +labs(title = "Observed genera through time", x= "Year of observation", y = "Number of individuals") + theme_bw() + theme(text=element_text(size=16), axis.text.x = element_text(colour = "grey20", angle = 90, size = 12))

```

## Last but not least Save your plot!

Put it in an object:
```{r}
combo_plot <- ggplot(data = yearly_sex_counts, mapping = aes( x = year, y=n, color= sex))+ geom_line() + facet_wrap(facets = vars(genus) ) +labs(title = "Observed genera through time", x= "Year of observation", y = "Number of individuals") + theme_bw() + theme(text=element_text(size=16), axis.text.x = element_text(colour = "grey20", angle = 90, size = 12))
```

Then save it...
```{r}
ggsave("combo_plot_final.png", combo_plot, width =10, dpi =300)

